Geometric Signs on Paleolithic Mobile Objects as Cultural Group Markers

### 2025-01-24 endofquarterpaper.qmd
editing

One of the most distinctive qualities of the Upper Paleolithic is the large rise in available artwork, both parietal and mobile. Archaeologists have frequently relied upon that artwork as a proxy for cultural difference in order to understand the grouping patterns of early modern humans. During this period, we also see a rise in organized settlements, more complex social organizations, and more advanced tool-making across Europe. Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca. 43,000 - 30,000 years BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by certain lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], many of which are decorated with geometric signs [@dutkiewicz2020].

We still lack a complete knowledge of the cultural diversity and movement patterns of the early modern humans from this era, despite a great deal of research into the subject. In this paper, we analyze the distribution and variation of geometric sign types found on mobile artefacts from the Upper Paleolithic in Europe during the Aurignacian technocomplex. The basis of this analysis is the idea that variations in signs can serve as an indicator of group membership and cultural similarity. This approach is supported by both ethnographic evidence throughout history, as well as similar archaeological studies done on ornamental and parietal art [@Pereira_Manen_2023; @Sauvet_2018; @Rigaud_dErrico_Vanhaeren_2015; @baker2024evidence; @Lycett_2019; @Vanhaeren_d]. In this study, we address the following questions: Are cultural groups evident in the distribution of geometric signs, and if so, what can they tell us about the cultural diversity and movement of Upper Paleolithic humans?

@@

Our understanding of the patterns of cultural and genetic diversity among the early modern humans of the Upper Paleolithic remains incomplete. Previous works have performed various cluster analysis methods upon variations in artefacts to represent and understand patterns of groups and cultural diversity. The studies have primarily relied upon personal ornamentation and parietal art, with few taking mobile objects into account.

Multivariate analyses of Gravettian beads from burial and occupation sites in Europe suggests an east to west cline of nine cultural groups [@baker2024evidence]. Seriation analysis, PCoA, and network analysis are used to identify the groups. The geographically discrete clusters identified encompass the regions of eastern, northwestern and central Europe, the northern and southern Iberian peninsula, southern and northern Italy, and the eastern and Western Mediterranean regions.

Similar analyses performed upon bead work patterns from the Aurignacian technocomplex as an indicator for linguistic diversity identify fourteen geographically cohesive clusters [@Vanhaeren_d]. Specifically, the groups are identified using seriation, correspondence and GIS analyses. The clusters identified are geographically discrete and sweep counter-clockwise throughout western France, northern Spain, the Pyrenees, and the Mediterranean region.

>>>>

~One of the most~ A distinctive quality~ies~ of the Upper Paleolithic is the large rise in ~available~ artwork, both parietal and mobile. Archaeologists have frequently relied upon that artwork as a proxy for cultural difference in order to understand the grouping patterns of early modern humans <!--BM: scholarly citations needed here-->. During this period, we also see a rise in organized settlements, more complex social organizations, and more advanced tool-making across Europe <!--BM: scholarly citations needed here-->. Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca. 43,000 - 30,000 years BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by certain lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], many of which are decorated with geometric signs [@dutkiewicz2020].

We still lack a ~complete knowledge~ <!--BM: I previously mentioned, this is not a compelling motivation, nothing can ever have a 'complete knowledge', so this is not a unique condition for our topic. Look at how other papers identify the gap in knowledge and follow their example and replace this sentence --> of the cultural diversity and movement patterns of the early modern humans from this era, despite a great deal of research into the subject. In this paper, we analyze the distribution and variation of geometric sign types found on mobile artefacts from the Upper Paleolithic in Europe during the Aurignacian ~technocomplex~ period. The basis of this analysis is the idea that variations in signs can serve as an indicator of group membership and cultural similarity. This approach is supported by both ethnographic evidence ~throughout history~ <!--BM: put scholarly citations on ethnography here-->, as well as similar archaeological studies ~done~ of~n~ ornamental and parietal art <!--BM: put scholarly citations on archaeology here--> [@Pereira_Manen_2023; @Sauvet_2018; @Rigaud_dErrico_Vanhaeren_2015; @baker2024evidence; @Lycett_2019; @Vanhaeren_d]. In this study, we address the following questions: Are cultural groups evident in the distribution of portable geometric signs, and if so, what can they tell us about the cultural diversity and movement of Upper Paleolithic humans?

@@

Our understanding of the patterns of cultural and genetic diversity among the early modern humans of the Upper Paleolithic ~remains incomplete~ <!--BM: as noted above and previously, not a strong opening, replace -->. Previous works have performed various <!--BM: avoid this weasel word 'various' it gives the reader no information --> cluster analysis methods upon variations in artefacts to represent and understand patterns of groups and cultural diversity. The studies <!--BM: not informative, this first paragraph is better replaced by a single sentence that very specifically introduces the previous work that the next paragraphs with describe in more detail: Gravettian beads, Aurignacian beads, quadrilateral geometric parietal signs  --> have primarily relied upon personal ornamentation and parietal art, with few taking mobile objects into account.

Multivariate analyses of Gravettian beads from burial and occupation sites in Europe suggests an east to west cline of nine cultural groups [@baker2024evidence]. Seriation analysis, PCoA, and network analysis ~are~ were <!--BM: everything in this background section needs to be in past tense, they have already done their work and it is complete --> used to identify the groups. The geographically discrete clusters identified encompass the regions of eastern, northwestern and central Europe, the northern and southern Iberian peninsula, southern and northern Italy, and the eastern and Western Mediterranean regions. <!--BM: this is a 3 sentence paragraph, that's a sign that something is wrong. Either add more detail, or combine with another para. I think adding more detail is ok here -->

Similar analyses performed upon bead work patterns from the Aurignacian ~technocomplex~ period as an indicator for linguistic diversity identify fourteen geographically cohesive clusters [@Vanhaeren_d]. Specifically, the groups ~are~ were identified using seriation, correspondence and GIS analyses. The clusters identified are geographically discrete and sweep counter-clockwise throughout western France, northern Spain, the Pyrenees, and the Mediterranean region.
